---
title: 'Teva memory - T cells - data preparation'
author: "Sarah"
date:  "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: 
    fig_crop: no
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
editor_options:
  
  chunk_output_type: console
---


## 1. Selection of cells from T cells clusters
```{r message=FALSE, warning=FALSE, cache=FALSE}
library(dplyr)
library(Seurat)
library(patchwork)
library(tidyverse)
library(sleepwalk)
library(gridExtra)
library(future)
library(reshape2)
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, 
                      message = FALSE, cache.lazy = FALSE)
setwd("//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups")


merged = readRDS("//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups/Objects/Teva_merged_object.rds")

T_cells_0= subset(x=merged, idents= c("0" = "0_CD8+ T cells", 
                                      "1" = "1_CD8+ T cells", 
                                      "7" = "7_CD8+ T cells + tumor cells",
                                      "8" = "8_CD4+ T cells"))
#dim(T_cells_0)
```

## 2. Normalize and scale the data
```{r message=FALSE, warning=FALSE, cache= TRUE}
# Normalize data
T_cells_0 = NormalizeData(T_cells_0)
T_cells_0 = FindVariableFeatures(T_cells_0, selection.method = "vst", nfeatures = 2000)
T_cells_0 = ScaleData(T_cells_0, vars.to.regress="percent.mt")

# Identify the 10 most highly variable genes
#top10_T_0 = head(VariableFeatures(T_cells_0), 10)

setwd("//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups/Figures/T cells/")
```

## 3. PCA
```{r message=FALSE, warning=FALSE, cache= TRUE,fig.width=6 , fig.height= 4}
# Perform linear dimensional reduction
T_cells_0 = RunPCA(T_cells_0, features = VariableFeatures(object = T_cells_0))

# Examine and visualize PCA results
DimPlot(T_cells_0, reduction = "pca", group.by="Treatment", cols = c("#CFBAE1", "#5C7D9D","#A4DEF9"), pt.size=1.1)

# Determine the ‘dimensionality’ of the dataset
ElbowPlot(T_cells_0, ndims = 40) + theme_classic()
ggsave(file = "plots/T/Elbow plot_0.png", dpi=300, width=6, height=6)

T_0_intermediate = saveRDS(T_cells_0, file="//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups/Objects/Teva_T_0_intermediate_object.rds")
```

6, 7, 9, 10, 18

Calibration of dimensions and resolution (in another file):
 --> 9 dimensions, resolution 0.5
 
 take-off cluster 11 (tumor cells)
 

## 4. Cluster the cells
```{r message=FALSE, warning=FALSE, cache= TRUE}
T_cells_0 = FindNeighbors(T_cells_0, dims = 1:9)
T_cells_0 = FindClusters(T_cells_0, resolution = 0.5)
```

## 5. Run non-linear dimensional reduction (UMAP/tSNE)
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=10 , fig.height= 10}
#T_cells_0 = RunUMAP(T_cells_0, dims = 1:9)
#DimPlot(T_cells_0, reduction = "umap", label= TRUE, pt.size=0.5)
#ggsave(file = "UMAP_0.png", dpi=300, width=12, height=12)

T_cells_0 = RunTSNE(T_cells_0, dims = 1:9, verbose = F)
DimPlot(T_cells_0, reduction = "tsne", label= TRUE, pt.size=0.5)
ggsave(file = "plots/T/tSNE_0.png", dpi=300, width=12, height=12)

TSNEPlot(T_cells_0, group.by ="Treatment", cols= c("#CFBAE1", "#5C7D9D","#A4DEF9"), pt.size=0.8)
ggsave(file = "plots/T/tSNE_0 - by treatment.png", dpi=300, width=12, height=12)
```


## 6. Remove cluster of tumor and non-T cells
```{r message=FALSE, warning=FALSE, cache=FALSE, fig.width=16 , fig.height= 24}
# T cells
FeaturePlot(T_cells_0, features = c("Ptprc", "Cd3e", "Cd8a", "Cd4", "Foxp3", "Mki67"), order=TRUE,pt.size=.5, reduction="tsne", ncol=3)
ggsave(file = "plots/T/features_T_0.png", dpi=300, width=10, height=6.67)

# Remove clusters 11 (non-T cells)
T_cells= subset(x=T_cells_0, idents= c(0,1,2,3,4,5,6,7,9,10))
saveRDS(T_cells, file="//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups/Objects/Teva_T_intermediate_object.rds")
```


## 7. Normalize and scale the data
```{r message=FALSE, warning=FALSE, cache= TRUE}
# Normalize data
T_cells = NormalizeData(T_cells)
T_cells = FindVariableFeatures(T_cells, selection.method = "vst", nfeatures = 2000)
T_cells = ScaleData(T_cells, vars.to.regress="percent.mt")

# Identify the 10 most highly variable genes
# top10_T = head(VariableFeatures(T_cells), 10)
```

## 8. PCA
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=6, fig.height= 6}
# Perform linear dimensional reduction
T_cells = RunPCA(T_cells, features = VariableFeatures(object = T_cells))

# Examine and visualize PCA results
DimPlot(T_cells, reduction = "pca", group.by="Treatment", cols = c("#CFBAE1", "#5C7D9D","#A4DEF9"),pt.size=1.1)

# Determine the ‘dimensionality’ of the dataset
ElbowPlot(T_cells, ndims = 40) + theme_classic()
ggsave(file = "plots/T/Elbow plot_T.png", dpi=300, width=6, height=6)
saveRDS(T_cells, file="//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups/Objects/T cells/Teva_T_3groups_before dim_object.rds")
```

Calibration of dimensions and resolution (in another file):
 --> 11 dimensions, resolution 0.3 
 
 
## 9. Cluster the cells
```{r message=FALSE, warning=FALSE, cache= TRUE}
T_cells  = FindNeighbors(T_cells , dims = 1:11)
T_cells  = FindClusters(T_cells , resolution = 0.3)
```


## 10. Run non-linear dimensional reduction (UMAP/tSNE)
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=10, fig.height= 10}
#T_cells  = RunUMAP(T_cells , dims = 1:10)
#DimPlot(T_cells , reduction = "umap", label= TRUE, pt.size=0.5)
#ggsave(file = "UMAP_T.png", dpi=300, width=12, height=12)

T_cells  = RunTSNE(T_cells , dims = 1:11, verbose = F)
DimPlot(T_cells , reduction = "tsne", label= TRUE, pt.size=0.5)
ggsave(file = "plots/T/tSNE_T.png", dpi=300, width=10, height=10)

TSNEPlot(T_cells, group.by ="Treatment", cols= c("#CFBAE1", "#5C7D9D","#A4DEF9"), pt.size=0.5)+
  theme(legend.text=element_text(size=20))
ggsave(file = "plots/T/tSNE - T - by treatment for teva .png", dpi=300, width=10, height=10)
```


## 11. Finding differentially expressed features (cluster biomarkers)
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=12, fig.height= 28}
T_cells.markers = FindAllMarkers(T_cells, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, return.thresh = 0.05)

Top100Markers_T =
  T_cells.markers %>%
  group_by(cluster) %>%
  top_n(n = 100, wt = avg_log2FC) %>%
  as.data.frame %>% 
  arrange(cluster,-avg_log2FC)

write_csv(Top100Markers_T, "//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups/Excels/memory_T cells_DE genes.csv")

T_cells.markers %>%
    group_by(cluster) %>%
    top_n(n = 10, wt = avg_log2FC) -> top10_T
DoHeatmap(T_cells, features = top10_T$gene) + NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "plots/T/T cells markers - heatmap.png", dpi=300, width=12, height=20)
saveRDS(T_cells, file="//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups/Objects/T cells/Teva_T_intermediate_object.rds")
```
