---
title: 'Teva memory - T cells - visualization'
author: "Michael"
date:  "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document: 
    fig_crop: no
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
editor_options:
  
  chunk_output_type: console
---

## 1. tSNE plot
```{r message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height= 8}
library(dplyr)
library(Seurat)
library(patchwork)
library(tidyverse)
library(sleepwalk)
library(gridExtra)
library(future)
library(reshape2)

setwd("//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups")


T_cells = readRDS(file="//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups/Objects/T cells/Teva_T_intermediate_object.rds")
 
T_cells  = RunTSNE(T_cells , dims = 1:10, verbose = F)
DimPlot(T_cells , reduction = "tsne", label= TRUE, pt.size=0.5)

TSNEPlot(T_cells, group.by ="Treatment", cols= c("#CFBAE1", "#5C7D9D","#A4DEF9"), pt.size=0.5)+
  theme(legend.text=element_text(size=20))
```


## 2. Subpopulations identification
### a. T cells populations
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=16, fig.height= 16}
FeaturePlot(T_cells, features = c("Cd3e", "Cd8a", "Cd4", "Foxp3"), order=TRUE,pt.size=1, reduction="tsne", ncol=2,label= TRUE)
ggsave(file = "Plots/T/features_T.png", dpi=300, width=12, height=12)
```

NKT markers
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=16, fig.height= 16}
FeaturePlot(T_cells, features = c("Klrb1", "Itga2", "Cxcr3", "Trdc", "Gzma", "Itgam",  "Gzmb", "Ncr1", "Gzmk"), order=TRUE,pt.size=1, reduction="tsne", ncol=3,label= TRUE)
ggsave(file = "Plots/T/features_NKT.png", dpi=300, width=20, height=20)
```

```{r eval=FALSE, fig.height=16, fig.width=16, message=FALSE, warning=FALSE, cache=TRUE}
FeaturePlot(T_cells, features = c("Itga2", "Trdc", "Gzma", "Itgam",  "Gzmb", "Ncr1"), order=TRUE,pt.size=1, reduction="tsne", ncol=3,label= TRUE)
ggsave(file = "Plots/T/features_NKT for teva.png", dpi=300, width=18, height=12)
```

γδ markers
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=16, fig.height= 16}
FeaturePlot(T_cells, features = c("Cxcr3", "Xcl1", "Ccl5", "Ctla2a", "Fcer1g", "Xcl1", "Ccl5", "Pglyrp1", "Cd160", "Klre1", "Fcer1g", "Klrc1", "Klrc2", "Klrd1", "Adam8", "Ccr6", "Ccr2", "Il1r1", "Il2ra", "Ltb4r1", 'Serpinb1a', 'Il23r', "Nr1d1", "Il17f", "Trdv4"), order=TRUE,pt.size=1, reduction="tsne", ncol=5, label= TRUE)
ggsave(file = "Plots/T/features_gdT.png", dpi=300, width=20, height=20)

FeaturePlot(T_cells, features = c("Ccl5", "Xcl1", "Klrd1", "Klrc1", "Klrc2", "Cd2", "Lck", "Itga4"), order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/features_gdT T1.png", dpi=300, width=20, height=20)
            
FeaturePlot(T_cells, features = c("Cxcr6", "Ccr2", "Il18
                                  r1", "Il7r", "Cd44", "Il17f", "Ccr6", "Icos", "Il1r1"), order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/features_gdT T17.png", dpi=300, width=20, height=20)
```


### b. activation/exhaustion markers
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 32}
# activation / exhaustion markers
FeaturePlot(T_cells, features = c("Pdcd1", "Havcr2", "Lag3", "Tigit", "Tnfrsf4", "Tnfrsf9", "Icos", "Ctla4", "Klrd1", "Irf8", "Klrc1", "Tox"), order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/features_T_activation.png", dpi=300, width=30, height=40)

table(T_cells$Treatment)

VlnPlot(T_cells, features = c('Havcr2'), flip= TRUE, split.by = "Treatment",pt.size = 0)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust=1, size = 15), axis.title.x = element_blank())+
   scale_fill_manual(values= c("#CFBAE1", "#5C7D9D","#A4DEF9"))+
geom_boxplot(alpha=0.3,show.legend = FALSE, outlier.shape = NA)+
        ylim(0,4)
ggsave(file = "Plots/T/Havcr2-3.png", dpi=300, width=8, height=4)

ateno_saline <- subset(T_cells, subset = Treatment == c("Saline", "αPD1-Il2"))
table(ateno_saline$Treatment) #M# making sure

VlnPlot(ateno_saline, features = c('Havcr2'), flip= TRUE, split.by = "Treatment",pt.size = 0)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())+
   scale_fill_manual(values= c("#CFBAE1","#A4DEF9"))+
geom_boxplot(alpha=0.3,show.legend = FALSE, outlier.shape = NA)
ggsave(file = "Plots/T/Havcr2-ateno_saline.png", dpi=300, width=8, height=4)
  
###
library(ggpubr)
# sub_t1 <- subset(ateno_saline, ident = as.character(cluster_id), subset = Treatment == "Saline")
# sub_t2 <- subset(ateno_saline, ident = as.character(cluster_id), subset = Treatment == "αPD1-Il2")
# expression_groupA <- sub_t1@assays$RNA@counts['Havcr2', ]
# expression_groupB <- sub_t2@assays$RNA@counts['Havcr2', ]
# ttest_result <- t.test(expression_groupA, expression_groupB)

p_values <- c()
for(cluster_id in levels(ateno_saline@active.ident)) {
  cluster_cells1 <- subset(ateno_saline, ident = as.character(cluster_id), subset = Treatment == "Saline")
  cluster_cells2 <- subset(ateno_saline, ident = as.character(cluster_id), subset = Treatment == "αPD1-Il2")
  expression_groupA <- cluster_cells1@assays$RNA@counts['Havcr2', ]
  expression_groupB <- cluster_cells2@assays$RNA@counts['Havcr2', ]
  p_values <- c(p_values, ttest_result$p.value)
}

p_values_bonferroni <- p.adjust(p_values, method = "bonferroni")

p_value_asterisks_bonferroni <- ifelse(p_values_bonferroni < 0.001, "***",
                                       ifelse(p_values_bonferroni < 0.01, "**",
                                              ifelse(p_values_bonferroni < 0.05, "*", "")))

x_levels <- seq(levels(ateno_saline@active.ident))

VlnPlot(ateno_saline, features = c('Havcr2'), flip= TRUE, split.by = "Treatment",pt.size = 0)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())+
   scale_fill_manual(values= c("#CFBAE1","#A4DEF9"))+
   geom_boxplot(alpha=0.3,show.legend = FALSE, outlier.shape = NA)+
   annotate("text", x = x_levels, y=3.1, label = p_value_asterisks_bonferroni, size = 6)+
        ylim(0,4)
ggsave(file = "Plots/T/Havcr2-ateno_saline_with_sig_test.png", dpi=300, width=8, height=4)




 VlnPlot(ateno_saline, features = c('Havcr2'), flip= TRUE, split.by = "Treatment",pt.size = 0)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())+
   scale_fill_manual(values= c("#CFBAE1","#A4DEF9"))+
   geom_boxplot(alpha=0.3,show.legend = FALSE, outlier.shape = NA)+
   annotate("text", x = x_levels, y=3.1, label = p_value_asterisks_bonferroni, size = 6)+
        ylim(0,4)+
  geom_label(data = data.frame(x_levels), aes(x = x_levels, y = 3.5, label = "{"),
             size = 6, angle = 90, vjust = -0.5, inherit.aes = FALSE)

```

### b.2(for sarah)
```{r}
VlnPlot(T_cells, features = c('Pdcd1'), flip= TRUE, split.by = "Treatment",pt.size = 0)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())+
   scale_fill_manual(values= c("#CFBAE1", "#5C7D9D","#A4DEF9"))+
geom_boxplot(alpha=0.3,show.legend = FALSE, outlier.shape = NA)+
        ylim(0,4)
ggsave(file = "Plots/T/Pdcd1-3.png", dpi=300, width=8, height=4)

library(stats)
# anova
data_for_anova <- T_cells@assays$RNA@data %>%
  as.data.frame() %>%
  select('Pdcd1', 'Treatment')

anova_result <- aov(Pdcd1 ~ Treatment, data = data_for_anova)

# tukey
tukey_result <- TukeyHSD(anova_result)

summary(anova_result)
print(tukey_result)




# two groups
ateno_saline <- subset(T_cells, subset = Treatment == c("Saline", "αPD1-Il2"))
table(ateno_saline$Treatment) #M# making sure

VlnPlot(ateno_saline, features = c('Pdcd1'), flip= TRUE, split.by = "Treatment",pt.size = 0)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())+
   scale_fill_manual(values= c("#CFBAE1","#A4DEF9"))+
geom_boxplot(alpha=0.3,show.legend = FALSE, outlier.shape = NA)+
        ylim(0,4)
ggsave(file = "Plots/T/Pdcd1-ateno_saline.png", dpi=300, width=8, height=4)

###
library(ggpubr)
# sub_t1 <- subset(ateno_saline, ident = as.character(cluster_id), subset = Treatment == "Saline")
# sub_t2 <- subset(ateno_saline, ident = as.character(cluster_id), subset = Treatment == "αPD1-Il2")
# expression_groupA <- sub_t1@assays$RNA@counts['Havcr2', ]
# expression_groupB <- sub_t2@assays$RNA@counts['Havcr2', ]
# ttest_result <- t.test(expression_groupA, expression_groupB)

p_values <- c()
for(cluster_id in levels(ateno_saline@active.ident)) {
  cluster_cells1 <- subset(ateno_saline, ident = as.character(cluster_id), subset = Treatment == "Saline")
  cluster_cells2 <- subset(ateno_saline, ident = as.character(cluster_id), subset = Treatment == "αPD1-Il2")
  expression_groupA <- cluster_cells1@assays$RNA@counts['Pdcd1', ]
  expression_groupB <- cluster_cells2@assays$RNA@counts['Pdcd1', ]
  p_values <- c(p_values, ttest_result$p.value)
}

p_values_bonferroni <- p.adjust(p_values, method = "bonferroni")

p_value_asterisks_bonferroni <- ifelse(p_values_bonferroni < 0.001, "***",
                                       ifelse(p_values_bonferroni < 0.01, "**",
                                              ifelse(p_values_bonferroni < 0.05, "*", "")))


x_levels <- seq(levels(ateno_saline@active.ident))

VlnPlot(ateno_saline, features = c('Pdcd1'), flip= TRUE, split.by = "Treatment",pt.size = 0)+
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())+
   scale_fill_manual(values= c("#CFBAE1","#A4DEF9"))+
   geom_boxplot(alpha=0.3,show.legend = FALSE, outlier.shape = NA)+
   annotate("text", x = x_levels, y=3.5, label = p_value_asterisks_bonferroni, size = 6)+
        ylim(0,4)
ggsave(file = "Plots/T/Pdcd1-ateno_saline_with_sig_test.png", dpi=300, width=8, height=4)

```


```{r eval=FALSE, fig.height=24, fig.width=24, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE}
# activation / exhaustion markers
FeaturePlot(T_cells, features = c("Pdcd1", "Havcr2", "Lag3", "Tigit", "Tnfrsf4", "Tnfrsf9", "Icos", "Ctla4", "Klrd1", "Irf8", "Klrc1", "Tox"), order=TRUE,pt.size=1, reduction="tsne", ncol=4, label= TRUE)
ggsave(file = "Plots/T/features_T_activation for teva.png", dpi=300, width=24, height=18)
```


```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 24}
### effector markers 
FeaturePlot(T_cells, features = c("Gzmb", "Prf1", "Ifng", "Mki67", "Gzmc", "Gzma", "Klrg1", "Il7r", "Cd52"),order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/features_T_effector.png", dpi=300, width=30, height=20)
```
Havcr2 = Tim3  
Tnfrsf4 = OX40  
Tnfrsf9 = 4-1BB  
Irf8: activation  
Klrc1 = NKG2A, checkpoint   
Tox and Nr4a2: involved in T cells exhaustion  

KLRG1, together with IL-7Ra: markers of effector t cells  
CD52: effector molecule  

```{r eval=FALSE, fig.height=24, fig.width=24, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE}
### effector markers 
FeaturePlot(T_cells, features = c("Gzmb", "Prf1", "Ifng", "Mki67", "Gzmc", "Gzma", "Klrg1"),order=TRUE,pt.size=1, reduction="tsne", ncol=4, label= TRUE)
ggsave(file = "Plots/T/features_T_effector for teva.png", dpi=300, width=24, height=12)
```


### c. naive-like (naive or memory cells)
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=16, fig.height= 16}
# naive
FeaturePlot(T_cells, features = c("Sell", "Il7r", "Ccr7", "Ms4a4c"),order=TRUE,pt.size=1, reduction="tsne", ncol=2, label= TRUE)
ggsave(file = "Plots/T/features_T_naive.png", dpi=300, width=20, height=20)
```

```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 16}
# memory 
FeaturePlot(T_cells, features = c("Sell", "Cd44", "Il7r", "Ccr7", "Tcf7", "Cxcr6"),order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/features_T_memory.png", dpi=300, width=30, height=20)
```

### d. effector memory markers
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 32}
FeaturePlot(T_cells, features = c("Bach2", "Prdm1", "Il7r", "Cd44", "Cx3cr1", "Id2","Klf2","Klrg1","Lef1","Stat4", "Tbx21", "Zeb2"),order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/features_T_effector memory.png", dpi=300, width=30, height=40)
```

```{r eval=FALSE, fig.height=24, fig.width=24, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE}
FeaturePlot(T_cells, features = c("Bach2", "Prdm1", "Il7r", "Cd44", "Cx3cr1", "Id2","Klf2","Klrg1","Lef1","Stat4", "Tbx21", "Zeb2"),order=TRUE,pt.size=1, reduction="tsne", ncol=4, label= TRUE)
ggsave(file = "Plots/T/features_T_effector memory for teva.png", dpi=300, width=32, height=24)
```

### e. central memory markers
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 40}
FeaturePlot(T_cells, features = c("Tcf7","Bach2", "Bcl16", "Ccr7", "Il7r","Cd27", "Cd44", "Sell","Id3", "Klf2", "Lcr7", "Lef1","Slamf6", "Stat3","Tfap4"),order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/features_T_central memory.png", dpi=300, width=30, height=50, limitsize=FALSE)
```


### f. resident memory markers
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=32, fig.height= 48}
FeaturePlot(T_cells, features = c("Prdm1", "Itgae", "Il7r", "Cd44","Itga1", "Cd69", "Crtam","Dusp6", "Egr2","Fosb", "Foxo3", "Gzma", "Gzmb","Znf683", "Id3", "Il2ra", "Irf4","Jun", "Junb", "Myc", "Nr4a1", "Rgs1"), order=TRUE, pt.size=1, reduction="tsne", ncol=4, label= TRUE)
ggsave(file = "Plots/T/features_T_resident memory.png", dpi=300, width=40, height=60, limitsize=FALSE)
```

### g. B cells
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 24}
FeaturePlot(T_cells, features = c("Cd19", "Ms4a1", "Iglc1", "Ighm", "Igkc"), order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/Features_T_B cells.png", dpi=300, width=30, height=20)
```

### h. markers from nature communication
https://www.nature.com/articles/s41467-021-23324-4

#### i. early activated
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 64}
FeaturePlot(T_cells, features = c("Gzmk", "Fos", "Cd69", "Zfp36", "Fosb", "CCl5", "Grzmm", "Dusp2", "Lyar", "Samd3", "Cxcr4", "Ctsw", "Anxa1", "Klrg1", "Aoah", "Tagap", "Klrd1", "Ier2", "Gzma", "Cst7", "Itm2c", "Parp8", "Btg2"),order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/features_T_nature_early activated.png", dpi=300, width=30, height=80, limitsize= FALSE)
```

#### ii. precursor exhausted
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 64}
FeaturePlot(T_cells, features = c("Lag3", "Xcl1", "Crtam", "Ifng", "Ccl4", "Pdcd1", "Dusp4", "Zeb2", "Nr4a2", "Sla", "Nkg7", "Tigit", "Ctsw", "Tnfrsf9", "Tox", "Lyst", "Tnfsf4", "Ccl3", "Gzmb", "Rab27a", "Prf1", "Cd70", "Plscr1", "Cxcl13"),order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/features_T_nature_precursor exhausted.png", dpi=300, width=30, height=80, limitsize= FALSE)
```
  
#### iii. terminally exhausted effector
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 64}
FeaturePlot(T_cells, features = c("Lag3", "Prf1", "Havcr2", "Gzmb", "Nkg7", "Ctsd", "Klrd1", "Id2", "Cst7", "Pdcd1", "Tnfrsf9", "Tigit", "Ctsw", "Ccl4", "Cd63", "Ccl3", "Ifng", "Cxcr6", "Fasl", "Rbpj", "Chst12", "Fam3c", "Csf1"),order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/features_T_nature_exhausted.png", dpi=300, width=30, height=80, limitsize= FALSE)
```
  
  
#### iv. naive-like (naive or central memory)
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 24}
FeaturePlot(T_cells, features = c("Ccr7", "Il7r", "Sell", "Tcf7", "Txk", "S1pr1", "Lef1", "Satb1"),order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/T_features_T_nature_naive-like.png", dpi=300, width=30, height=30)
```

  
#### v. effector memory
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height= 64}
FeaturePlot(T_cells, features = c("Gzma", "Gzmk", "Nkg7", "Ctsw", "Gzmb", "Ccl5", "Cst7", "Prf1", "Abi3", "Fasl", "Itm2c", "1500009L16Rik", "Eomes", "Chst12", "Ccr5", "Hcst", "Aoah", "Hopx", "Slamf7", "Cxcr3", "Oasl1", "F2r", "Cxcr6"),order=TRUE,pt.size=1, reduction="tsne", ncol=3, label= TRUE)
ggsave(file = "Plots/T/T_features_T_nature_effector memory.png", dpi=300, width=30, height=80, limitsize= FALSE)
```


## 3. Heatmap 

### a. activation/exhaustion/effector
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=6}
DoHeatmap(T_cells, features = c("Pdcd1", "Havcr2", "Lag3", "Tigit", "Tnfrsf4", "Tnfrsf9", "Icos", "Ctla4", "Klrd1", "Irf8", "Klrc1", "Tox1", "Cx3cr1", "Xcl1", "Gzmb", "Prf1", "Mki67", "Ifng", "Gzmc", "Gzma", "Klrg1", "Il7r", "Cd52"))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_activated effector.png", dpi=300, width=30, height=30)
```

### b. naive/memory
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=6}
DoHeatmap(T_cells, features = c("Sell", "Il7r", "Ccr7", "Ms4a4c","Sell", "Cd44", "Il7r", "Ccr7", "Tcf7", "Cxcr6", "Txk", "S1pr1", "Lef1", "Satb1" ))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_naive memory.png", dpi=300, width=30, height=30)
```

### c. effector memory
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=6}
DoHeatmap(T_cells, features = c("Bach2", "Prdm1", "Il7r", "Cd44", "Cx3cr1", "Id2", "Klf2", "Klrg1", "Lef1", "Stat4","Tbx21","Zeb2", "Gzma", "Gzmk", "Nkg7", "Cd8a", "Cd8b1", "Ctsw", "Gzmb", "Ccl5", "Cst7", "Prf1", "Abi3", "Fasl", "Itm2c", "1500009L16Rik", "Eomes", "Chst12", "Ccr5", "Hcst", "Aoah", "Hopx", "Slamf7", "Cxcr3", "Oasl1", "F2r", "Cxcr6"))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_effector memory.png", dpi=300, width=30, height=30)
```

### d. central memory
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=6}
DoHeatmap(T_cells, features = c("Bach2", "Bcl16", "Ccr7", "Il7r","Cd27", "Cd44", "Sell","Id3", "Klf2", "Lcr7", "Lef1","Slamf6", "Stat3", "Tcf7","Tfap4"))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_central memory.png", dpi=300, width=30, height=30)
```

### e. resident memory
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=6}
DoHeatmap(T_cells, features = c("Prdm1", "Itgae", "Il7r", "Cd44", "Itga1", "Cd69", "Crtam", "Cx3cr1", "Dusp6", "Egr2", "Fabp5", "Fos", "Fosb", "Foxo3", "Gzma", "Gzmb", "Znf683", "Id3", "Il2ra", "Irf4", "Jun", "Junb", "Myc", "Nr4a1", "Rgs1"))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_resident memory.png", dpi=300, width=30, height=30)
```

Cd244a = 2B4  
Phenotypically, the expression of inhibitory receptors increased gradually, including those of PD-1, LAG3, Tim-3, CD160, 2B4, and CTLA-4.57 In terms of transcription factors, TCF-1, TOX, T-bet, Eomes, and FOXO1, which are related to the exhaustion and memory formation, they all undergo changes, respectively.

### f. nature communication markers
#### i. early activated
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=6}
DoHeatmap(T_cells, features = c("Gzmk", "Fos", "Cd69", "Zfp36", "Fosb", "CCl5", "Grzmm", "Dusp2", "Lyar", "Samd3", "Cxcr4", "Ctsw", "Anxa1", "Klrg1", "Aoah", "Tagap", "Klrd1", "Ier2", "Gzma", "Cst7", "Itm2c", "Parp8", "Btg2"))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_nature_early activated.png", dpi=300, width=30, height=30)
```
  
#### ii. precursor exhausted
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=6}
DoHeatmap(T_cells, features = c("Lag3", "Xcl1", "Crtam", "Ifng", "Ccl4", "Pdcd1", "Dusp4", "Zeb2", "Nr4a2", "Sla", "Nkg7", "Tigit", "Ctsw", "Tnfrsf9", "Tox", "Lyst", "Tnfsf4", "Ccl3", "Gzmb", "Rab27a", "Prf1", "Cd70", "Plscr1", "Cxcl13"))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_nature_precursor exhausted.png", dpi=300, width=30, height=30)
```

#### iii. terminally exhausted effector
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=6}
DoHeatmap(T_cells, features = c("Lag3", "Prf1", "Havcr2", "Gzmb", "Nkg7", "Ctsd", "Klrd1", "Id2", "Cst7", "Pdcd1", "Tnfrsf9", "Tigit", "Ctsw", "Ccl4", "Cd63", "Ccl3", "Ifng", "Cxcr6", "Fasl", "Rbpj", "Chst12", "Fam3c", "Csf1"))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_nature_exhausted.png", dpi=300, width=30, height=30)
```

#### iv. naive-like
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=6}
DoHeatmap(T_cells, features = c("Ccr7", "Il7r", "Sell", "Tcf7", "Txk", "S1pr1", "Lef1", "Satb1"))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_nature_naive-like.png", dpi=300, width=30, height=30)
```

#### v. effector memory
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=6}
DoHeatmap(T_cells, features = c("Gzma", "Gzmk", "Nkg7", "Ctsw", "Gzmb", "Ccl5", "Cst7", "Prf1", "Abi3", "Fasl", "Itm2c", "1500009L16Rik", "Eomes", "Chst12", "Ccr5", "Hcst", "Aoah", "Hopx", "Slamf7", "Cxcr3", "Oasl1", "F2r", "Cxcr6"))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_nature_effector memory.png", dpi=300, width=30, height=30)
```


## 4. Annotation
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=25, fig.height= 20}
T_cells@active.ident = T_cells$seurat_clusters
T_cells = RenameIdents(T_cells, 
                       "0" = "0_Effector CD8+", 
                       "1" = "1_Effector memory CD8+", 
                       "2" = "2_Naive? (/memory?) CD8+", 
                       "3" = "3_Activated/effector? CD8+", 
                       "4" = "4_Tregs and CD4+ helper", 
                       "5" = "5_Proliferating CD8+",
                       "6" = "6_Activated (?) CD8+")
DimPlot(T_cells, reduction = "tsne", label=TRUE, label.size= 7, pt.size=2)+
  theme(legend.text= element_text(size=24))
ggsave(file = "Plots/T/T cells - annotated tsne - with names.png", dpi=300, width=25, height=20)

DimPlot(T_cells, reduction = "tsne", label=FALSE, pt.size=2)+
  theme(legend.text= element_text(size=24))
ggsave(file = "Plots/T/T cells - annotated tsne.png", dpi=300, width=25, height=20)
saveRDS(T_cells, file = "//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups/Objects/T cells/T cells_object.rds")

T_cells <- readRDS("//asafmadilabfs/asafmadilab$/michael/Madi lab/scRNA seq Pipeline/m/sara_teva/3_groups/Objects/T cells/T cells_object.rds")

```


## 5. Contribution of each group
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=10, fig.height= 6}
ggplot(T_cells@meta.data, aes(x=T_cells@active.ident, fill=Treatment)) + geom_bar(position="fill")+
    scale_fill_manual(values= c("#CFBAE1", "#5C7D9D","#A4DEF9"))+
    theme_classic()+
    labs(x="")+
    theme(legend.title=element_text(size=18),legend.text= element_text(size=14), 
          axis.text.x=element_text(size=8,angle=45, hjust=1))
ggsave(file = "Plots/T/T_stacked bar plot.png", dpi=300, width=10, height=10)
```

```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=18, fig.height= 10}
ggplot(T_cells@meta.data, aes(x=T_cells@active.ident, fill=Treatment)) + geom_bar(stat="count",position=position_dodge())+
    scale_fill_manual(values= c("#CFBAE1", "#5C7D9D","#A4DEF9"))+
    theme_classic()+
    labs(x="")+
    theme(legend.title=element_text(size=18),legend.text= element_text(size=14), 
          axis.text.x=element_text(size=12,angle=45, hjust=1))

ggsave(file = "Plots/T/T_bar plot.png", dpi=300, width=15, height=10)
```

```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=18, fig.height= 10}
a=prop.table(table(Idents(T_cells), T_cells$Treatment), margin = 2)
b=as.data.frame.matrix(a) 
b$cluster=rownames(b)
proportions=melt(b)
colnames(proportions) = c("cluster", "Treatment", "prop")

proportions$cluster = factor (proportions$cluster, levels=c("0_Effector CD8+", "1_Effector memory CD8+",  "2_Naive? (/memory?) CD8+", "3_Activated/effector? CD8+", "4_Tregs and CD4+ helper", "5_Proliferating CD8+","6_Activated (?) CD8+"))

ggplot(proportions, aes(x=cluster, y=prop, fill=Treatment))+
  geom_bar(stat="identity",position=position_dodge())+
    theme_classic()+
    labs(x="", y="% of total cells")+
    theme(legend.title=element_text(size=18),legend.text= element_text(size=14), 
          axis.text.x=element_text(size=12,angle=45, hjust=1))
ggsave(file = "Plots/T/T_bar plot_proportions.png", dpi=300, width=15, height=10)       
```


## 6. Specific genes/pathways

### a. Batf

```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=6, fig.height=6}
FeaturePlot(T_cells, features = c("Batf"),pt.size=1, reduction="tsne")
ggsave(file = "Plots/T/features_T_Batf.png", dpi=300, width=10, height=10)
```


```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=18, fig.height=12}
library(ggpubr)
library(tidyr)
VlnPlot(T_cells, features = c("Batf"))+stat_compare_means()
ggsave(file = "Plots/T/Vln_T_Batf.png", dpi=300, width=15, height=10)
```

### b. glycolysis

visualization of glycolysis pathway in different ways:
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=24, fig.height=24}
FeaturePlot(T_cells, features = c("Pfkm", 'Psmc4', 'Nup85', 'Ppha1', 'Cdk1', "Cenpa", "Gpi", "Nup155", "Gpc1"),order=TRUE,pt.size=1, reduction="tsne", ncol=3)
ggsave(file = "Plots/T/features_T_glycolysis.png", dpi=300, width=30, height=30)
```

```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=10, fig.height=6}
VlnPlot(T_cells, features = c('Psmc4', 'Ppha1', 'Cdk1', "Cenpa", "Gpi"), stack=TRUE, flip= TRUE, split.by = "seurat_clusters")+theme_classic()+theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())
ggsave(file = "Plots/T/Vln_T_glycolysis.png", dpi=300, width=15, height=10)

VlnPlot(T_cells, features = c('Psmc4', 'Ppha1', 'Cdk1', "Cenpa", "Gpi"), stack=TRUE, flip= TRUE, split.by = "Treatment")+
     scale_fill_manual(values= c("#CFBAE1", "#5C7D9D","#A4DEF9"))+
  theme_classic()+theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())

```

```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=10, fig.height=5}
DotPlot(object=T_cells, features = c('Psmc4', 'Ppha1', 'Cdk1', "Cenpa")) + coord_flip()+theme_classic()+theme(axis.text.x = element_text(angle = 45, hjust=1))
ggsave(file = "Plots/T/dot_T_glycolysis.png", dpi=300, width=10, height=7)
```


```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=8}
glycolysis_list = list(c("Pfkm", 'Psmc4', 'Nup85', 'Ppha1', 'Cdk1', "Cenpa", "Gpi", "Nup155", "Gpc1"))
T_cells = AddModuleScore(object = T_cells, features = glycolysis_list, name = "glycolysis_signature")
FeaturePlot(object = T_cells, features = "glycolysis_signature1", reduction = "tsne")
ggsave(file = "Plots/T/signature_T_glycolysis.png", dpi=300, width=10, height=10)
```

more markers:
```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=40, fig.height=72}
FeaturePlot(T_cells, features = c("Hk2","Aldoc","Pfkl","Nup85","Nup155", "Tpi1","Adpgk","Pfkfb1","Pfkfb4","Hk3", "Pfkfb2","Pfkfb3","Nup35","Ndc1","Gnpda2", "Aldoa","Pgm2l1","Pkm","Nup93","Pfkm","Nup54",'Nup37', 'Gpi1', 'Aaas',"Nup205",'Bpgm','Nup133',"Nup43","Nup88",'Nupl2','Nup160',"Gnpda1","Nup188","Nup107","Pom121","Gapdh","Eno1b","Eno3","Gapdhs","Pgk1","Eno1",'Nup98',"Seh1l","Nup62","Aldoa"), order=TRUE,pt.size=1, reduction="tsne", ncol=5)
ggsave(file = "Plots/T/feature_T_glycolysis_2.png", dpi=300, width=50, height=90, limitsize=FALSE)
```

```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=8}
glycolysis_list2 = list(c("Hk2","Aldoc","Pfkl","Nup85","Nup155", "Tpi1","Adpgk","Pfkfb1","Pfkfb4","Hk3", "Pfkfb2","Pfkfb3","Nup35","Ndc1","Gnpda2", "Aldoa","Pgm2l1","Pkm","Nup93","Pfkm","Nup54",'Nup37', 'Gpi1', 'Aaas',"Nup205",'Bpgm','Nup133',"Nup43","Nup88",'Nupl2','Nup160',"Gnpda1","Nup188","Nup107","Pom121","Gapdh","Eno1b","Eno3","Gapdhs","Pgk1","Eno1",'Nup98',"Seh1l","Nup62","Aldoa"))
T_cells = AddModuleScore(object = T_cells, features = glycolysis_list2, name = "glycolysis2_signature")
FeaturePlot(object = T_cells, features = "glycolysis2_signature1", reduction = "tsne")
ggsave(file = "Plots/T/signature_T_glycolysis2.png", dpi=300, width=10, height=10)
```

```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=8, fig.height=8}
DoHeatmap(T_cells, features = c("Pfkm", 'Psmc4', 'Nup85', 'Ppha1', 'Cdk1', "Cenpa", "Gpi", "Nup155", "Gpc1", "Hk2","Aldoc","Pfkl","Nup85","Nup155", "Tpi1","Adpgk","Pfkfb1","Pfkfb4","Hk3", "Pfkfb2","Pfkfb3","Nup35","Ndc1","Gnpda2", "Aldoa","Pgm2l1","Pkm","Nup93","Pfkm","Nup54",'Nup37', 'Gpi1', 'Aaas',"Nup205",'Bpgm','Nup133',"Nup43","Nup88",'Nupl2','Nup160',"Gnpda1","Nup188","Nup107","Pom121","Gapdh","Eno1b","Eno3","Gapdhs","Pgk1","Eno1",'Nup98',"Seh1l","Nup62","Aldoa"))+ NoLegend() + scale_fill_gradientn(colors = c("blue", "white", "red"))
ggsave(file = "Plots/T/T_heatmap_glycolysis.png", dpi=300, width=20, height=30)
```


### c. oxydative phosphorylation
```{r eval=FALSE, fig.height=8, fig.width=8, message=FALSE, warning=FALSE, cache=TRUE, include=FALSE}
FeaturePlot(T_cells, features = c("Abcd1","Afg1l","Ak4","Atp7a","Atpsckmt","Bid",	"Ccnb1",	"Cdk1",	"Chchd10","Coa6",	"Coq7",	"Coq9", "Cox4i2", "Cox7a1",	"Cyc1",	"Cycs",	"Dguok",	"Dld","Dnajc30",	"Fxn",	"Gadd45gip1",	"Iscu",	"Msh2",	"mt-Atp6",	"mt-Atp8",	"mt-Co1", "mt-Co2",	"mt-Co3",	"mt-Cytb",	"mt-Nd1",	"mt-Nd2",	"mt-Nd3",	"mt-Nd4",	"mt-Nd4l",	"mt-Nd5",	"mt-Nd6",	"Mtch2",	"Myc", "Ndufa1", "Ndufa10",	"Ndufa11"	,"Ndufa12",	"Ndufa13",	"Ndufa2",	"Ndufa3",	"Ndufa5",	"Ndufa6",	"Ndufa7",	"Ndufa8",	"Ndufa9",	"Ndufab1",	"Ndufaf1", "Ndufc1",	"Ndufs1",	"Ndufs2","Ndufv1",	"Nipsnap2",	"Nupr1", "Pde12", "Pde2a",	"Pink1",	"Ppif",	"Sdha",	"Sdhaf2", "Shmt2",	"Slc25a23",	"Slc25a33",	"Stoml2","Tefm",	"Tmem135",	"Tnf",	"Uqcc2",	"Uqcc3",	"Uqcr10" ),order=TRUE,pt.size=1, reduction="tsne", ncol=9)                    
ggsave(file = "Plots/T/features_T_oxidative phosphorylation.png", dpi=300, width=90, height=90, limitsize=FALSE)
```

```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=10, fig.height=15}
VlnPlot(T_cells, features = c("Ccnb1",	"Cdk1",	"Chchd10","Coa6",	"Cyc1",	"Cycs",	"Dguok",	"Dld",	"Fxn"), stack=TRUE, flip= TRUE, split.by = "seurat_clusters")+theme_classic()+theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())
ggsave(file = "Plots/T/Vln_T_oxidative phosphorylation_1.png", dpi=300, width=20, height=30, limitsize=FALSE)
                              
VlnPlot(T_cells, features = c("Gadd45gip1",	"Iscu","mt-Atp6",	"mt-Atp8",	"mt-Co1", "mt-Co2",	"mt-Co3",	"mt-Cytb",	"mt-Nd1",	"mt-Nd2"), stack=TRUE, flip= TRUE, split.by = "seurat_clusters")+theme_classic()+theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())
ggsave(file = "Plots/T/Vln_T_oxidative phosphorylation_2.png", dpi=300, width=20, height=30, limitsize=FALSE)
                              
VlnPlot(T_cells, features = c("mt-Nd3",	"mt-Nd4",	"mt-Nd4l",	"mt-Nd5","Mtch2", "Ndufa1", "Ndufa10",	"Ndufa11"	,"Ndufa12"), stack=TRUE, flip= TRUE, split.by = "seurat_clusters")+theme_classic()+theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())
ggsave(file = "Plots/T/Vln_T_oxidative phosphorylation_3.png", dpi=300, width=20, height=30, limitsize=FALSE)
        
VlnPlot(T_cells, features = c("Ndufa13",	"Ndufa2",	"Ndufa3",	"Ndufa5",	"Ndufa6",	"Ndufa7",	"Ndufa8",	"Ndufa9",	"Ndufab1"), stack=TRUE, flip= TRUE, split.by = "seurat_clusters")+theme_classic()+theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())
ggsave(file = "Plots/T/Vln_T_oxidative phosphorylation_4.png", dpi=300, width=20, height=30, limitsize=FALSE)

VlnPlot(T_cells, features = c("Ndufa10",	"Ndufa11"	,"Ndufa12", "Ndufc1",	"Ndufs1",	"Ndufs2","Ndufv1", "Pde2a",	"Sdha",	"Sdhaf2"), stack=TRUE, flip= TRUE, split.by = "seurat_clusters")+theme_classic()+theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())
ggsave(file = "Plots/T/Vln_T_oxidative phosphorylation_5.png", dpi=300, width=20, height=30, limitsize=FALSE)
```

```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=10, fig.height=10}
VlnPlot(T_cells, features = c("Stoml2",	"Tmem135",	"Tnf",	"Uqcc2",	"Uqcc3",	"Uqcr10"), stack=TRUE, flip= TRUE, split.by = "seurat_clusters")+theme_classic()+theme(axis.text.x = element_text(angle = 45, hjust=1), axis.title.x = element_blank())
ggsave(file = "Plots/T/Vln_T_oxidative phosphorylation_6.png", dpi=300, width=20, height=25, limitsize=FALSE)            
```


```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=10, fig.height=15}
DotPlot(object=T_cells, features = c("Abcd1","Afg1l","Ak4","Atp7a","Atpsckmt","Bid","Ccnb1","Cdk1","Chchd10","Coa6",	"Coq7","Coq9","Cox4i2","Cox7a1","Cyc1","Cycs","Dguok","Dld","Dnajc30","Fxn","Gadd45gip1","Iscu","Msh2","mt-Atp6","mt-Atp8","mt-Co1","mt-Co2","mt-Co3","mt-Cytb","mt-Nd1","mt-Nd2","mt-Nd3","mt-Nd4","mt-Nd4l","mt-Nd5","mt-Nd6","Mtch2","Myc","Ndufa1","Ndufa10","Ndufa11","Ndufa12","Ndufa13","Ndufa2","Ndufa3","Ndufa5","Ndufa6","Ndufa7","Ndufa8","Ndufa9","Ndufab1","Ndufaf1","Ndufc1","Ndufs1","Ndufs2","Ndufv1","Nipsnap2","Nupr1","Pde12","Pde2a","Pink1","Ppif","Sdha","Sdhaf2","Shmt2","Slc25a23","Slc25a33","Stoml2","Tefm","Tmem135","Tnf","Uqcc2","Uqcc3","Uqcr10")) + coord_flip()+theme_classic()+theme(axis.text.x = element_text(angle = 45, hjust=1))
ggsave(file = "Plots/T/dot_T_oxidative phosphorylation.png", dpi=300, width=15, height=19)

```


```{r message=FALSE, warning=FALSE, cache= TRUE, fig.width=6, fig.height=6}
oxidation_list = list(c("Abcd1","Afg1l","Ak4","Atp7a","Atpsckmt","Bid","Ccnb1","Cdk1","Chchd10","Coa6","Coq7","Coq9","Cox4i2","Cox7a1","Cyc1","Cycs","Dguok","Dld","Dnajc30","Fxn","Gadd45gip1","Iscu","Msh2","mt-Atp6","mt-Atp8","mt-Co1","mt-Co2","mt-Co3","mt-Cytb","mt-Nd1","mt-Nd2","mt-Nd3","mt-Nd4","mt-Nd4l","mt-Nd5","mt-Nd6","Mtch2","Myc","Ndufa1","Ndufa10","Ndufa11","Ndufa12","Ndufa13","Ndufa2","Ndufa3","Ndufa5","Ndufa6","Ndufa7","Ndufa8","Ndufa9",	"Ndufab1",	"Ndufaf1", "Ndufc1",	"Ndufs1",	"Ndufs2", "Ndufv1",	"Nipsnap2",	"Nupr1", "Pde12", "Pde2a",	"Pink1",	"Ppif", "Sdha","Sdhaf2","Shmt2", "Slc25a23","Slc25a33", "Stoml2","Tefm", "Tmem135","Tnf","Uqcc2", "Uqcc3","Uqcr10"))
T_cells = AddModuleScore(object = T_cells, features = oxidation_list, name = "oxidation_signature")
FeaturePlot(object = T_cells, features = "oxidation_signature1", reduction = "tsne")
ggsave(file = "Plots/T/signature_T_oxidation.png", dpi=300, width=10, height=10)
```

&nbsp;
<hr />
<p style="text-align: center;">Work by Michael Monsonego</a></p>
<p style="text-align: center;"> <a href="https://www.asafmadilab.com/">Madi Lab</a></p>
&nbsp;
